---
title: "Hands-on Exercise 9a"
subtitle: "Choropleth Mapping with R"
author: "Yee Weicong Mark"
date: "June 15, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    css: styles.css
---

## [21 Choropleth Mapping with R]{style="color:  #4682B4; font-size: 38px;"}

## [21.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colours. A social scientist may need a choropleth map to portray the spatial distribution of the aged population of Singapore by Master Plan 2014 Subzone Boundary.

This chapter plots functional and truthful choropleth maps with an R package called tmap.

## [21.2 Getting Started]{style="color:  #2C5282; font-size: 28px;"}

The key R package for this exercise is tmap. Four other packages are used alongside it.

1.  **readr** for importing delimited text files,
2.  **tidyr** for tidying data,
3.  **dplyr** for wrangling data,
4.  **sf** for handling geospatial data.

readr, tidyr and dplyr are part of tidyverse, so only tidyverse needs to be installed.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## [21.3 Importing Data into R]{style="color:  #2C5282; font-size: 28px;"}

## [21.3.1 The Data]{style="color:  #366092; font-size: 24px;"}

Two datasets are used to create the choropleth map.

1.  Master Plan 2014 Subzone Boundary (Web) (`MP14_SUBZONE_WEB_PL`) in ESRI shapefile format. This geospatial data holds the geographical boundary of Singapore at the planning subzone level, based on the Urban Redevelopment Authority (URA) Master Plan 2014.
2.  Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011 to 2020 (`respopagesextod2011to2020.csv`). This aspatial data file carries no coordinates, but its PA and SZ fields serve as unique identifiers to geocode to the `MP14_SUBZONE_WEB_PL` shapefile.

## [21.3.2 Importing Geospatial Data into R]{style="color:  #366092; font-size: 24px;"}

`st_read()` of sf imports the `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_WEB_PL")
```

The content of `mpsz` is examined with the code chunk below.

```{r}
mpsz
```

::: callout-note
## Things to Learn from the Code Chunk Above

Only the first ten records are displayed. The simple feature data frame holds 323 features and 15 fields, with a MULTIPOLYGON geometry projected in SVY21.
:::

## [21.3.3 Importing Attribute Data into R]{style="color:  #366092; font-size: 24px;"}

`read_csv()` of readr imports `respopagesextod2011to2020.csv` into an R data frame called `popdata`.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## [21.3.4 Data Preparation]{style="color:  #366092; font-size: 24px;"}

A data table with year 2020 values is required before a thematic map can be prepared. The table should hold the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL and DEPENDENCY.

1.  YOUNG covers age group 0 to 4 until age group 20 to 24,
2.  ECONOMY ACTIVE covers age group 25 to 29 until age group 60 to 64,
3.  AGED covers age group 65 and above,
4.  TOTAL covers all age groups,
5.  DEPENDENCY is the ratio between the young and aged groups against the economy-active group.

### [21.3.4.1 Data Wrangling]{style="color:  #366092; font-size: 24px;"}

The wrangling uses `pivot_wider()` of tidyr and `mutate()`, `filter()`, `group_by()` and `select()` of dplyr.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG,
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
         rowSums(.[13:15]))%>%
  mutate(`AGED`=rowSums(.[16:21])) %>%
  mutate(`TOTAL`=rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
         /`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
         `ECONOMY ACTIVE`, `AGED`,
         `TOTAL`, `DEPENDENCY`)
```

### [21.3.4.2 Joining the Attribute Data and Geospatial Data]{style="color:  #366092; font-size: 24px;"}

The values in the PA and SZ fields are first converted to uppercase. The PA and SZ fields mix upper and lowercase, while SUBZONE_N and PLN_AREA_N are uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

`left_join()` of dplyr joins the geographical data and attribute table using the planning subzone name, with SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
## Things to Learn from the Code Chunk Above

`left_join()` is called with the `mpsz` simple feature data frame as the left table so the output stays a simple feature data frame.
:::

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## [21.4 Choropleth Mapping Geospatial Data Using tmap]{style="color:  #2C5282; font-size: 28px;"}

Two approaches prepare a thematic map with tmap. The first plots a thematic map quickly with `qtm()`. The second plots a highly customisable thematic map with tmap elements.

## [21.4.1 Plotting a Choropleth Map Quickly by Using qtm()]{style="color:  #366092; font-size: 24px;"}

`qtm()` is the quickest way to draw a choropleth map with tmap. It is concise and gives a good default visualisation in many cases.

::: panel-tabset
## Plot

```{r}
#| echo: false
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

## Code

```{r}
#| eval: false
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

`tmap_mode("plot")` produces a static map. The "view" option produces an interactive map instead. The `fill` argument maps the DEPENDENCY attribute to colour.
:::

## [21.4.2 Creating a Choropleth Map by Using tmap's Elements]{style="color:  #366092; font-size: 24px;"}

`qtm()` makes the aesthetics of individual layers harder to control. tmap's drawing elements produce a high-quality cartographic choropleth map.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```
:::

### [21.4.2.1 Drawing a Base Map]{style="color:  #366092; font-size: 24px;"}

The basic building block of tmap is `tm_shape()` followed by one or more layer elements such as `tm_fill()` and `tm_polygons()`. `tm_shape()` defines the input data and `tm_polygons()` draws the planning subzone polygons.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons()
```
:::

### [21.4.2.2 Drawing a Choropleth Map Using tm_polygons()]{style="color:  #366092; font-size: 24px;"}

Assigning the target variable to `tm_polygons()` draws a choropleth map of that variable by planning subzone.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

The default interval binning is "pretty". The default colour scheme is YlOrRd of ColorBrewer. Missing values are shaded grey by default.
:::

### [21.4.2.3 Drawing a Choropleth Map Using tm_fill() and tm_borders()]{style="color:  #366092; font-size: 24px;"}

`tm_polygons()` is a wrapper of `tm_fill()` and `tm_borders()`. `tm_fill()` shades the polygons with the default colour scheme and `tm_borders()` adds the shapefile borders. The code chunk below draws a choropleth map with `tm_fill()` alone.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```
:::

The subzones are shaded by their dependency values. `tm_borders()` adds the subzone boundaries.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

The `alpha` argument sets transparency between 0 (fully transparent) and 1 (fully opaque), defaulting to the colour's own alpha. `tm_borders()` also takes `col` for border colour, `lwd` for border line width (default 1) and `lty` for border line type (default "solid").
:::

## [21.4.3 Data Classification Methods of tmap]{style="color:  #366092; font-size: 24px;"}

Most choropleth maps employ a data classification method. Classification groups a large number of observations into data ranges or classes. tmap provides ten methods, namely fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher and jenks. The `style` argument of `tm_fill()` or `tm_polygons()` selects the method.

### [21.4.3.1 Plotting Choropleth Maps with Built-in Classification Methods]{style="color:  #366092; font-size: 24px;"}

The code chunk below uses the jenks method with five classes.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

The code chunk below uses the equal method.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

::: callout-tip
## Reading the Plot

The quantile classification spreads observations more evenly across classes than the equal classification. The choice of method changes the visual story, so the same data can read very differently under different classifications.
:::

### [21.4.3.2 Plotting Choropleth Map with Custom Break]{style="color:  #366092; font-size: 24px;"}

For the built-in styles, the category breaks are computed internally. The `breaks` argument of `tm_fill()` overrides these defaults. The breaks include a minimum and maximum, so n categories require n+1 break elements in increasing order.

Descriptive statistics on the variable guide the break points. The code chunk below computes the statistics of the DEPENDENCY field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

The break points are set at 0.60, 0.70, 0.80 and 0.90, with a minimum of 0 and a maximum of 1.00. The breaks vector is therefore c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```
:::

**Do It Yourself**

The appearance of a choropleth map shifts with the classification, so testing alternatives is worthwhile.

1.  Using what you have learned, prepare choropleth maps with different classification methods supported by tmap and compare their differences.
2.  Prepare choropleth maps with the same classification method but different numbers of classes (2, 6, 10, 20). Compare the output maps. What observation can you draw?

**Answer to the First Task**

Four classification methods are applied to the same DEPENDENCY field with five classes each, then placed side by side with `tmap_arrange()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 8
dep_equal <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval")

dep_quantile <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")

dep_jenks <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks")

dep_sd <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "sd",
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Standard deviation")

tmap_arrange(dep_equal, dep_quantile,
             dep_jenks, dep_sd, ncol = 2)
```

## Code

```{r}
#| eval: false
dep_equal <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval")

dep_quantile <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")

dep_jenks <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks", n = 5,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks")

dep_sd <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "sd",
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Standard deviation")

tmap_arrange(dep_equal, dep_quantile,
             dep_jenks, dep_sd, ncol = 2)
```
:::

::: callout-tip
## Reading the Plot

Equal interval divides the range into bands of equal width. The dependency ratio is right-skewed with a few very high values, so most subzones fall into the lowest band and the map reads as almost one colour. Quantile places an equal count of subzones in each class, which spreads the colours across the map and exposes far more spatial variation. Jenks groups the values at their natural breaks and produces a balanced map close to the quantile result. Standard deviation shades each subzone by its distance from the mean and highlights the extreme outliers rather than the general pattern. The same data therefore tells four different stories, and quantile and jenks reveal the most usable spatial structure for this variable.
:::

**Answer to the Second Task**

The quantile method is held fixed while the number of classes is set to 2, 6, 10 and 20.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 8
dep_n2 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 2,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("2 classes")

dep_n6 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 6,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("6 classes")

dep_n10 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 10,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("10 classes")

dep_n20 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 20,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("20 classes")

tmap_arrange(dep_n2, dep_n6,
             dep_n10, dep_n20, ncol = 2)
```

## Code

```{r}
#| eval: false
dep_n2 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 2,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("2 classes")

dep_n6 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 6,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("6 classes")

dep_n10 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 10,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("10 classes")

dep_n20 <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile", n = 20,
        values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("20 classes")

tmap_arrange(dep_n2, dep_n6,
             dep_n10, dep_n20, ncol = 2)
```
:::

::: callout-tip
## Reading the Plot

With two classes the map shows only a broad split between low and high dependency and hides almost all local variation. Six and ten classes add progressively more detail and the spatial pattern becomes richer and easier to read. Twenty classes shade the map so finely that neighbouring colours are hard to tell apart and the legend grows too long to scan, so the map approaches a continuous wash that is harder to interpret. More classes show more detail only up to a point, and a moderate count of around five to seven balances detail against legibility for most readers.
:::

## [21.4.4 Colour Scheme]{style="color:  #366092; font-size: 24px;"}

tmap supports colour ramps defined by the user or drawn from the RColorBrewer package.

### [21.4.4.1 Using ColourBrewer Palette]{style="color:  #366092; font-size: 24px;"}

The preferred colour is assigned to the `values` argument of `tm_scale_intervals()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```
:::

A "-" prefix reverses the colour shading.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```
:::

## [21.4.5 Map Layouts]{style="color:  #366092; font-size: 24px;"}

Map layout refers to the combination of all map elements into a cohesive map. Map elements include the mapped objects, the title, the scale bar, the compass, the margins and the aspect ratios.

### [21.4.5.1 Map Legend]{style="color:  #366092; font-size: 24px;"}

Several `tm_legend()` options change the placement, format and appearance of the legend.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```
:::

### [21.4.5.2 Map Style]{style="color:  #366092; font-size: 24px;"}

tmap allows a wide variety of layout settings, called through `tmap_style()`. The code chunk below uses the classic style.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```
:::

### [21.4.5.3 Cartographic Furniture]{style="color:  #366092; font-size: 24px;"}

tmap provides arguments for other map furniture such as a compass, scale bar and grid lines. `tm_compass()`, `tm_scale_bar()` and `tm_grid()` add these elements to the choropleth map.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```
:::

The default style is reset with the code chunk below.

```{r}
tmap_style("white")
```

## [21.4.6 Drawing Small Multiple Choropleth Maps]{style="color:  #366092; font-size: 24px;"}

Small multiple maps, also called facet maps, arrange many maps side by side and sometimes stacked vertically. They show how spatial relationships change with respect to another variable such as time. tmap plots small multiple maps in three ways, by assigning multiple values to at least one aesthetic argument, by defining a group-by variable in `tm_facets()` and by creating multiple stand-alone maps with `tmap_arrange()`.

### [21.4.6.1 By Assigning Multiple Values to at Least One Aesthetic Argument]{style="color:  #366092; font-size: 24px;"}

In this example, small multiple choropleth maps are created by defining `ncols` in `tm_fill()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```
:::

Multiple values are assigned to more than one aesthetic argument in this example.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020)+
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"),
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020)+
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"),
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```
:::

### [21.4.6.2 By Defining a Group-by Variable in tm_facets()]{style="color:  #366092; font-size: 24px;"}

Multiple small choropleth maps are created with `tm_facets()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by="REGION_N",
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by="REGION_N",
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```
:::

### [21.4.6.3 By Creating Multiple Stand-alone Maps with tmap_arrange()]{style="color:  #366092; font-size: 24px;"}

Multiple small choropleth maps are created as stand-alone maps and combined with `tmap_arrange()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
youngmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## Code

```{r}
#| eval: false
youngmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```
:::

## [21.4.7 Mapping Spatial Object Meeting a Selection Criterion]{style="color:  #366092; font-size: 24px;"}

A selection function maps spatial objects meeting a criterion instead of creating small multiple maps.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

## Code

```{r}
#| eval: false
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
:::

## [21.5 Reference]{style="color:  #2C5282; font-size: 28px;"}

Kam, T. S. (2025). *R for Visual Analytics*, Chapter 21, Choropleth Mapping with R. Singapore Management University.

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 21 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: Cartographer** 🗺️

*Press any key to continue...*
:::

```{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
```
