---
title: "Hands-on Exercise 9c"
subtitle: "Analytical Mapping"
author: "Yee Weicong Mark"
date: "June 15, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    css: styles.css
---

## [23 Analytical Mapping]{style="color:  #4682B4; font-size: 38px;"}

## [23.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

## [23.1.1 Objectives]{style="color:  #366092; font-size: 24px;"}

This exercise gives hands-on experience using appropriate R methods to plot analytical maps.

## [23.1.2 Learning Outcome]{style="color:  #366092; font-size: 24px;"}

By the end of this exercise you will be able to use appropriate functions of tmap and tidyverse to:

1.  import geospatial data in rds format into the R environment,
2.  create cartographic quality choropleth maps with appropriate tmap functions,
3.  create a rate map,
4.  create a percentile map,
5.  create a box map.

## [23.2 Getting Started]{style="color:  #2C5282; font-size: 28px;"}

## [23.2.1 Installing and Loading Packages]{style="color:  #366092; font-size: 24px;"}

The sf, tmap and tidyverse packages are installed and loaded into the R environment.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

## [23.2.2 Importing Data]{style="color:  #366092; font-size: 24px;"}

A prepared dataset called `NGA_wp.rds` is used. It is a polygon feature data frame holding information on the water points of Nigeria at the Local Government Area (LGA) level. `read_rds()` imports the file into R.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## [23.3 Basic Choropleth Mapping]{style="color:  #2C5282; font-size: 28px;"}

## [23.3.1 Visualising Distribution of Non-Functional Water Point]{style="color:  #366092; font-size: 24px;"}

Two choropleth maps are built and arranged side by side. The first maps functional water points and the second maps total water points, both by LGA.

::: panel-tabset
## Plot

```{r}
#| echo: false
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of total water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```

## Code

```{r}
#| eval: false
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of total water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```
:::

## [23.4 Choropleth Map for Rates]{style="color:  #2C5282; font-size: 28px;"}

Mapping rates rather than counts matters because water points are not equally distributed in space. Without accounting for how many water points are present in each area, the map shows total water point size rather than the topic of interest.

## [23.4.1 Deriving Proportion of Functional and Non-Functional Water Points]{style="color:  #366092; font-size: 24px;"}

The proportion of functional and non-functional water points in each LGA is tabulated. `mutate()` of dplyr derives two fields, `pct_functional` and `pct_nonfunctional`.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

## [23.4.2 Plotting Map of Rate]{style="color:  #366092; font-size: 24px;"}

A choropleth map shows the distribution of the percentage of functional water points by LGA.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

## Code

```{r}
#| eval: false
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```
:::

## [23.5 Extreme Value Maps]{style="color:  #2C5282; font-size: 28px;"}

Extreme value maps are variations of common choropleth maps where the classification highlights extreme values at the lower and upper ends of the scale to identify outliers. They add spatial features to approaches used in non-spatial exploratory data analysis.

## [23.5.1 Percentile Map]{style="color:  #366092; font-size: 24px;"}

The percentile map is a special quantile map with six categories, 0 to 1 percent, 1 to 10 percent, 10 to 50 percent, 50 to 90 percent, 90 to 99 percent and 99 to 100 percent. The breakpoints are derived with the base R quantile command, passing an explicit vector of cumulative probabilities, c(0, .01, .1, .5, .9, .99, 1). The begin and endpoint must be included.

### [23.5.1.1 Data Preparation]{style="color:  #366092; font-size: 24px;"}

Step 1 excludes records with NA values.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2 creates the customised classification and extracts values.

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

::: callout-note
## Things to Learn from the Code Chunk Above

Extracting a variable from an sf data frame extracts the geometry as well. Many base R functions cannot handle the geometry, and `quantile()` gives an error. `st_set_geometry(NULL)` drops the geometry field.
:::

### [23.5.1.2 Why Write Functions]{style="color:  #366092; font-size: 24px;"}

Writing a function has three advantages over copy and paste. A function takes an evocative name that makes the code easier to understand. When requirements change, the code is updated in one place rather than many. The chance of incidental mistakes from copying and pasting is removed.

### [23.5.1.3 Creating the get.var Function]{style="color:  #366092; font-size: 24px;"}

An R function extracts a variable as a vector out of an sf data frame. The `vname` argument is the variable name as a quoted character, and `df` is the sf data frame. The function returns a vector of values without a column name.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### [23.5.1.4 A Percentile Mapping Function]{style="color:  #366092; font-size: 24px;"}

A percentile mapping function is written with the code chunk below.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_polygons(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle)
}
```

### [23.5.1.5 Test Drive the Percentile Mapping Function]{style="color:  #366092; font-size: 24px;"}

The function is run with the code chunk below.

::: panel-tabset
## Plot

```{r}
#| echo: false
percentmap("total_wp", NGA_wp)
```

## Code

```{r}
#| eval: false
percentmap("total_wp", NGA_wp)
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

This is a bare bones implementation. Additional arguments such as the title and legend positioning could be passed to customise the map.
:::

## [23.5.2 Box Map]{style="color:  #366092; font-size: 24px;"}

A box map is an augmented quartile map with an additional lower and upper category. When lower outliers are present, the starting break is the minimum value and the second break is the lower fence. When no lower outliers are present, the starting break is the lower fence and the second break is the minimum value, with no observations falling between them.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

## Code

```{r}
#| eval: false
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```
:::

A box map displays summary statistics on a choropleth map using the principles of the boxplot. A custom breaks specification is used, and the break points vary depending on whether lower or upper outliers are present.

### [23.5.2.1 Creating the boxbreaks Function]{style="color:  #366092; font-size: 24px;"}

The R function below creates break points for a box map. The `v` argument is the vector of observations and `mult` is the multiplier for the interquartile range (IQR), defaulting to 1.5. The function returns a vector of seven break points.

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

### [23.5.2.2 Creating the get.var Function]{style="color:  #366092; font-size: 24px;"}

The R function below extracts a variable as a vector out of an sf data frame. The `vname` argument is the variable name as a quoted character and `df` is the sf data frame.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### [23.5.2.3 Test Drive the Newly Created Function]{style="color:  #366092; font-size: 24px;"}

The new function is tested with the code chunk below.

```{r}
var <- get.var("wp_nonfunctional", NGA_wp)
boxbreaks(var)
```

### [23.5.2.4 Boxmap Function]{style="color:  #366092; font-size: 24px;"}

The R function below creates a box map. The `vnam` argument is the variable name as a quoted character, `df` is the simple features polygon layer, `legtitle` is the legend title, `mtitle` is the map title and `mult` is the multiplier for the IQR. The function returns a tmap element.

```{r}
boxmap <- function(vnam, df,
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier",
                     "< 25%",
                     "25% - 50%",
                     "50% - 75%",
                     "> 75%",
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle)
}
```

::: panel-tabset
## Plot

```{r}
#| echo: false
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

## Code

```{r}
#| eval: false
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
:::

::: callout-tip
## Reading the Plot

The box map separates the lower and upper outlier LGAs from the four quartile classes. The outlier categories draw the eye to areas whose non-functional water-point counts fall outside the bulk of the distribution.
:::

## [23.6 Reference]{style="color:  #2C5282; font-size: 28px;"}

Kam, T. S. (2025). *R for Visual Analytics*, Chapter 23, Analytical Mapping. Singapore Management University.

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 23 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: Outlier Hunter** 🎯

*Press any key to continue...*
:::

```{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
```
