---
title: "Hands-on Exercise 2"
author: "Yee Weicong Mark"
date: "April 18, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    css: styles.css
---

## [Beyond ggplot2 Fundamentals]{style="color:  #4682B4; font-size: 38px;"}

## [2.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

This chapter goes beyond basic `ggplot2` and introduces **four extension packages** that together can transform the exploratory plots into publication ready figures. The three main skills are:

1.  **Controlling label placement** - `ggrepel` solves the common problem of overlapping text labels by repelling annotations away from each other and their data points using a simple swap from `geom_text()` to `geom_text_repel()`

2.  **Professional publication themes** - `ggthemes` and `hrbrthemes` elevate plot aesthetics beyond the eight built-in themes, offering publication-style looks (*Economist*, *FiveThirtyEight*, *WSJ*) and typography-focused designs respectively, with fine-grained control over font sizes, grid-lines, and label placement.

3.  **Combining multiple plots into one figure** - `patchwork` addresses figure composition, using operator syntax (`+` for side-by-side, `/` for stacked, `()` for grouping, `&` to apply themes uniformly) to combine multiple plots into a single coherent figure, complete with auto-tagging for academic referencing and inset capabilities for layered visualisations.

These skills operate at three distinct layers (1) **readability** (annotations), (2) **aesthetics** (themes), and (3) **composition** (layout) and applying them together lets you move a plot from "looks fine in RStudio" to "professional look ready for a report or paper".

## [2.2 Getting Started]{style="color:  #2C5282; font-size: 28px;"}

## [2.2.1 Installing and Loading the Required Libraries]{style="color:  #366092; font-size: 24px;"}

Four ggplot2 extension packages are used alongside tidyverse:

| Package      | What it does                           |
|--------------|----------------------------------------|
| `ggrepel`    | Stops text labels from overlapping     |
| `ggthemes`   | Themes                                 |
| `hrbrthemes` | Typography-centric themes              |
| `patchwork`  | Combine multiple plots into one figure |

Code chunk below will be used to check if these packages have been installed and also will load them onto your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

The package ‘hrbrthemes’ is not available for this version of R, therefore hrbrthemes have to be installed from Github using the following:

`remotes::install_github("hrbrmstr/hrbrthemes")`

## [2.2.2 Importing Data]{style="color:  #366092; font-size: 24px;"}

The file *Exam_data* is used for this exercise. The file can be imported using the using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package.

`exam_data <- read_csv("chap02/data/Exam_data.csv")`

::: callout-warning
This is a **relative path**. It only works if your working directory is the parent of `chap02/`. Otherwise you will get `Error: 'chap02/data/Exam_data.csv' does not exist`.

You can check your working directory using `getwd().`
:::

For my case, I will leave it in the parent folder and my file will be imported using the following:

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## [2.3 Beyond ggplot2 Annotation: ggrepel]{style="color:  #2C5282; font-size: 28px;"}

**The problem:** With many data points, text labels overlap into an unreadable blob.

**The fix:** `ggrepel` pushes labels apart and draws connector lines back to their points.

Simple swap:

-   `geom_text()` → `geom_text_repel()`

<!-- -->

-   `geom_label()` → `geom_label_repel()`

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm, size=0.5) +
geom_label(aes(label = ID), hjust = .5, vjust = -.5) + coord_cartesian(xlim=c(0,100), ylim=c(0,100)) + ggtitle("English scores versus Maths scores for Primary 3")
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) + 
  geom_point() + 
  geom_smooth(method=lm, size=0.5) +
geom_label(aes(label = ID), 
           hjust = .5, 
           vjust = -.5) + 
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) + 
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## [2.3.1 Working with ggrepel]{style="color:  #366092; font-size: 24px;"}

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## [2.4 Beyond ggplot2 Themes]{style="color:  #2C5282; font-size: 28px;"}

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`. Below are a few examples of the different themes.

::: panel-tabset
## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

## gray

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

## classic

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_classic() +
  ggtitle("Distribution of Maths scores") 
```

## linedraw

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_linedraw() +
  ggtitle("Distribution of Maths scores") 
```

## void

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_void() +
  ggtitle("Distribution of Maths scores") 
```
:::

## [2.4.1 Working with ggtheme Package]{style="color:  #366092; font-size: 24px;"}

Themes mimicking famous publications: **Tufte, Economist, FiveThirtyEight, WSJ, Stata, Excel**. The example uses `theme_economist()` having a recognisable light-blue background, serif title. This helps to provide instant professional looks for reports.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
:::

## [2.4.2 Working with hrbthemes Package]{style="color:  #366092; font-size: 24px;"}

[**hrbrthemes**](https://cran.r-project.org/web/packages/hrbrthemes/index.html) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
:::

The second goal centers around productivity for a production workflow in terms of font choice, label spacing and readability. Consult [this vignette](Why%20hrbrthemes?%20•%20hrbrthemes) to learn more.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

::: callout-tip
## What can we learn from the code chunk above?

-   `axis_title_size` argument is used to increase the font size of the axis title to 18
-   `base_size` argument is used to increase the default axis label to 15
-   `grid` argument is used to remove the x-axis grid lines
:::

## [2.5 Beyond Single Graph]{style="color:  #2C5282; font-size: 28px;"}

Multiple plots together often tell a better story than one. This section builds three plots (histograms of MATHS, ENGLISH, then a MATHS-vs-ENGLISH scatterplot) then combines them.

::: panel-tabset
## Plot (Maths)

```{r}
#| label: plot-maths
#| echo: false
p1 <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100,
                 color = "grey25", fill = "grey90") + 
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of Maths scores")
p1
```

## Code (Maths)

```{r}
#| ref.label: plot-maths
#| eval: false
```

## Plot (English)

```{r}
#| label: plot-english
#| echo: false
p2 <- ggplot(data = exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20, boundary = 100,
                 color = "grey25", fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of English scores")
p2
```

## Code (English)

```{r}
#| ref.label: plot-english
#| eval: false
```

## Plot (Maths vs English)

```{r}
#| label: plot-scatter
#| echo: false
p3 <- ggplot(data = exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +  
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
p3
```

## Code (Maths vs English)

```{r}
#| ref.label: plot-scatter
#| eval: false
```
:::

## [2.5.1 Creating Composite Graphics: Patchwork Method]{style="color:  #366092; font-size: 24px;"}

There are different methods to combine plots such as (`gridExtra::grid.arrange()`, `cowplot::plot_grid()`), but `patchwork` has the cleanest syntax and the plots combine like arithmetic:

| Operator | Effect                            |
|----------|-----------------------------------|
| `+`      | Two columns layout (Side by side) |
| `/`      | Two rows layout (Stacked rows)    |
| `|`      | Side by side (explicit)           |
| `()`     | Create subplot group              |

## [2.5.2 Combining Two ggplot2 Graphs]{style="color:  #366092; font-size: 24px;"}

Figure in the tabset below shows a composite of two histograms created using patchwork. Note how simple the syntax used to create the plot!

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
p1 + p2
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
p1 + p2
```
:::

## [2.5.3 Combining Three ggplot2 Graphs]{style="color:  #366092; font-size: 24px;"}

We can plot more complex composite by using appropriate operators. For example, the composite figure below is plotted by using:

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

Example: `(p1 / p2) | p3` means stack p1 on p2 (left column), put p3 next to them (right column).

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
(p1 / p2) | p3
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
(p1 / p2) | p3
```
:::

To learn more about, refer to [Plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html).

## [2.5.4 Combining a Composite Figure with Tag]{style="color:  #366092; font-size: 24px;"}

In order to identify subplots in text, **patchwork** also provides auto-tagging capabilities.

## Basic tag levels

| Value | Output                   | Example              |
|-------|--------------------------|----------------------|
| `'A'` | Uppercase Latin letters  | A, B, C, D, E...     |
| `'a'` | Lowercase Latin letters  | a, b, c, d, e...     |
| `'1'` | Arabic numerals          | 1, 2, 3, 4, 5...     |
| `'I'` | Uppercase Roman numerals | I, II, III, IV, V... |
| `'i'` | Lowercase Roman numerals | i, ii, iii, iv, v... |

The figure below shows the uppercase roman numerals.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```
:::

## [2.5.5 Creating Figure with Insert]{style="color:  #366092; font-size: 24px;"}

Beside providing functions to place plots next to each other based on the provided layout. With [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) of **patchwork**, we can place one or several plots or graphic elements freely on top or below another plot. This can be used as a main plot with a zoomed in inset, or a summary overlaid on detail.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```
:::

::: callout-tip
## Coordinate gotcha

left, bottom, right, top are in relative units (0–1) of the main plot — not pixels, not data coordinates. Easy to accidentally place the inset off-screen. Start with safe values: inset_element(p_inset, left = 0.6, bottom = 0.6, right = 1, top = 1)
:::

## [2.5.6 Creating a Composite Figure by Using Patchwork and ggtheme]{style="color:  #366092; font-size: 24px;"}

Figure below is created by combining patchwork and theme_economist() of ggthemes package discussed earlier.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## Code

```{r}
#| eval: false
#| attr-source: "style='overflow-x: auto; white-space: pre;'"
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
:::

::: callout-tip
## Applying Theme to Partial Plots.

Critical **&** vs **+** distinction:

-   `+`adds to the last plot only
-   & adds to every plot in the composite"

(p1 + p2 + p3) & theme_economist() \# ✅ all three themed

(p1 + p2 + p3) + theme_economist() \# ❌ only p3 themed
:::

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 2 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: ggplot2 Ninja** 🥷

*Press any key to continue...*
:::

```{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
```