---
title: "Chapter 13: Creating Ternary Plot with R"
subtitle: "Hands-on Exercise 5"
author: "Yee Weicong Mark"
date: "May 14, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    css: styles.css
---

## [13 Creating Ternary Plot with R]{style="color:  #4682B4; font-size: 38px;"}

## [13.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

Ternary plots display three-part compositional data, meaning data where the parts must sum to a fixed total such as 1 or 100%. The display is an equilateral triangle with each side scaled from 0 to 1 and each vertex representing 100% of one component. This chapter covers static ternary plots with `ggtern` and interactive ones with `plotly`, then explores five design alternatives.

## [13.2 Installing and Launching R Packages]{style="color:  #2C5282; font-size: 28px;"}

Three R packages are required:

1. **ggtern**, a `ggplot2` extension specially designed for ternary diagrams
2. **plotly**, used here for its native `scatterternary` trace type which is preferred over `ggplotly()` conversion because the latter does not fully translate `ggtern` geometries
3. **tidyverse** for data import and shape transformation

**knitr** is also loaded so `kable()` can produce clean previews.

```{r}
pacman::p_load(plotly, ggtern, tidyverse, knitr)
```

::: callout-note
## Things to learn from the code chunk above

`ggtern` masks several `ggplot2` functions when loaded, because the package overrides standard ggplot internals to support the triangular coordinate system. If conflicts arise, explicit namespacing with `ggplot2::` or `ggtern::` resolves them.
:::

## [13.3 Data Preparation]{style="color:  #2C5282; font-size: 28px;"}

## [13.3.1 The Data]{style="color:  #366092; font-size: 24px;"}

The dataset `respopagsex2000to2018_tidy.csv` is from the Singapore Department of Statistics. It records resident population by planning area, subzone, age group, sex, and year from 2000 to 2018.

## [13.3.2 Importing Data]{style="color:  #366092; font-size: 24px;"}

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")

kable(head(pop_data))
```

## [13.3.3 Preparing the Data]{style="color:  #366092; font-size: 24px;"}

To plot a ternary diagram, the data must be reshaped from long to wide format and three compositional measures derived: **YOUNG** (under 25), **ACTIVE** (25 to 64), and **OLD** (65 and above).

::: callout-warning
## A bug in the R4VA recipe worth fixing

The R4VA tutorial derives the three measures using positional column indexing:

```r
mutate(YOUNG = rowSums(.[4:8])) %>%
mutate(ACTIVE = rowSums(.[9:16])) %>%
mutate(OLD = rowSums(.[17:21]))
```

This assumes the columns from `pivot_wider()` come out in natural age order, but they do not. The age bands in the source data are named `AGE0-4`, `AGE05-9`, `AGE10-14`, and so on. After `pivot_wider()`, the columns appear in **first-appearance order** rather than alphabetical or natural age order. Because `AGE05-9` first appears far down the source file, it ends up at column 13 instead of column 5.

Applied to the columns as they actually come out, the positional recipe produces:

1. **YOUNG (cols 4 to 8)** sums `AGE0-4` + `AGE10-14` + `AGE15-19` + `AGE20-24` + `AGE25-29`. Children aged 5 to 9 are missing, and 25-to-29-year-olds are mistakenly included.
2. **ACTIVE (cols 9 to 16)** sums `AGE30-34` through `AGE45-49`, then `AGE05-9` (column 13), then `AGE50-54` through `AGE60-64`. Children aged 5 to 9 are mixed into the working-age group, and 25-to-29-year-olds are missing.
3. **OLD (cols 17 to 21)** sums `AGE65-69` through `AGE85over`. This one is correct.

A quick verification: for Ang Mo Kio Town Centre in 2018, the tutorial recipe reports YOUNG = 1,360. Adding the actual under-25 bands gives 180 + 270 + 320 + 300 + 260 = 1,330, not 1,360. The 30-person discrepancy is `AGE25-29` (300) being wrongly included while `AGE05-9` (270) is excluded.

The fix is to select columns by name using `rowSums(across(c(...)))`. The backticks handle hyphens in the column names.
:::

```{r}
agpop_mutated <- pop_data %>%
  mutate(Year = as.character(Year)) %>%
  pivot_wider(names_from = AG, values_from = Population) %>%
  mutate(
    YOUNG = rowSums(across(c(`AGE0-4`, `AGE05-9`, `AGE10-14`, `AGE15-19`, `AGE20-24`))),
    ACTIVE = rowSums(across(c(`AGE25-29`, `AGE30-34`, `AGE35-39`, `AGE40-44`, `AGE45-49`,
                              `AGE50-54`, `AGE55-59`, `AGE60-64`))),
    OLD = rowSums(across(c(`AGE65-69`, `AGE70-74`, `AGE75-79`, `AGE80-84`, `AGE85over`))),
    TOTAL = YOUNG + ACTIVE + OLD
  ) %>%
  filter(Year == 2018) %>%
  filter(TOTAL > 0)

kable(head(agpop_mutated))
```

::: callout-note
## Things to learn from the code chunk above

`pivot_wider()` replaces the deprecated `spread()` function. It widens the dataset so each age band becomes its own column.

`rowSums(across(c(...)))` selects columns by name rather than by position. Backticks around the names handle the hyphens. `TOTAL = YOUNG + ACTIVE + OLD` is more transparent than the original `rowSums(.[22:24])` and cannot break if upstream columns are reordered.

`filter(Year == 2018)` restricts the analysis to a single cross-section, and `filter(TOTAL > 0)` removes subzones with zero population.
:::

## [13.4 Plotting Ternary Diagram with R]{style="color:  #2C5282; font-size: 28px;"}

## [13.4.1 Plotting a Static Ternary Diagram]{style="color:  #366092; font-size: 24px;"}

The `ggtern()` function works like `ggplot()` but expects three positional aesthetics: `x`, `y`, and `z`, mapping to the three vertices of the triangle.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point()
```

## Code

```{r}
#| eval: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point()
```
:::

::: callout-note
## Things to learn from the code chunk above

`ggtern()` automatically constructs the triangular coordinate system. The three aesthetics `x`, `y`, and `z` map to the three vertices, and the package internally normalises each row so the three values sum to 1. Most standard ggplot2 geoms have ternary equivalents.
:::

::: callout-tip
## Reading the plot

Each point is one Singapore subzone in 2018. The cluster sits closer to the ACTIVE vertex (right side) than to YOUNG or OLD, which means most subzones are dominated by working-age residents. A small group of points drifts towards the OLD vertex, indicating subzones with disproportionately ageing populations. Subzones near the YOUNG vertex are rarer in 2018, consistent with the overall demographic transition.

This is the analytical payoff of a ternary plot. A scatterplot of YOUNG against OLD would lose the ACTIVE dimension, and a parallel coordinates plot would lose the immediate visual sense of which subzones lean towards which extreme.
:::

## [13.4.2 Adding Title and Theme]{style="color:  #366092; font-size: 24px;"}

The default theme can be replaced with `theme_rgbw()` for cleaner gridlines and a title added via `labs()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point() +
  labs(title = "Singapore population structure by subzone, 2018",
       subtitle = "Each point is one subzone; vertices represent 100% of the named age group") +
  theme_rgbw()
```

## Code

```{r}
#| eval: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point() +
  labs(title = "Singapore population structure by subzone, 2018",
       subtitle = "Each point is one subzone; vertices represent 100% of the named age group") +
  theme_rgbw()
```
:::

::: callout-note
## Things to learn from the code chunk above

`theme_rgbw()` is one of several `ggtern` themes (others include `theme_bw()`, `theme_classic()`, `theme_dark()`).
:::

## [13.4.3 Plotting an Interactive Ternary Diagram]{style="color:  #366092; font-size: 24px;"}

`plot_ly()` supports a native `scatterternary` trace type, which is preferred over `ggplotly(ggtern_plot)` because the latter does not faithfully translate `ggtern` geometry.

```{r}
# Reusable function for annotation
label <- function(txt) {
  list(
    text = txt,
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper",
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# Reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt,
    tickformat = ".0%",
    tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"),
  baxis = axis("Active"),
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated,
  a = ~YOUNG,
  b = ~ACTIVE,
  c = ~OLD,
  color = I("black"),
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Singapore subzones, 2018"),
    ternary = ternaryAxes
  )
```

::: callout-note
## Things to learn from the code chunk above

`plot_ly()` uses `a`, `b`, `c` (not `x`, `y`, `z`) for ternary aesthetics, which is a quirk of the underlying plotly.js API. `tickformat = ".0%"` formats the axis ticks as percentages with no decimals. The `label()` and `axis()` helpers are reusable, so defining them once avoids repetition if multiple ternary plots are produced.
:::

::: callout-tip
## Reading the plot

The interactive version supports zoom, pan, and hover-to-identify. For a dataset with hundreds of subzones, this turns a dense static cluster into an explorable map. Hovering over an outlier subzone (one nearest the OLD vertex, for instance) reveals the subzone name directly, rather than guessing from position alone.
:::

## [13.5 Exploring Alternatives]{style="color:  #2C5282; font-size: 28px;"}

Some variations below to enhance the plot.

## [13.5.1 Alternative 1: Colouring by Planning Area]{style="color:  #366092; font-size: 24px;"}

Adding colour reveals whether subzones cluster geographically.

::: panel-tabset
## Plot

```{r}
#| echo: false
agpop_top <- agpop_mutated %>%
  group_by(PA) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  filter(PA %in% c("Bedok", "Tampines", "Jurong West", "Woodlands", "Sengkang", "Punggol"))

ggtern(data = agpop_top,
       aes(x = YOUNG, y = ACTIVE, z = OLD, colour = PA)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Population structure across selected planning areas, 2018",
       subtitle = "Newer towns (Punggol, Sengkang) skew young; older towns (Bedok) skew old",
       colour = "Planning Area") +
  theme_rgbw()
```

## Code

```{r}
#| eval: false
agpop_top <- agpop_mutated %>%
  group_by(PA) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  filter(PA %in% c("Bedok", "Tampines", "Jurong West", "Woodlands", "Sengkang", "Punggol"))

ggtern(data = agpop_top,
       aes(x = YOUNG, y = ACTIVE, z = OLD, colour = PA)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Population structure across selected planning areas, 2018",
       subtitle = "Newer towns (Punggol, Sengkang) skew young; older towns (Bedok) skew old",
       colour = "Planning Area") +
  theme_rgbw()
```
:::

::: callout-note
## Things to learn from the code chunk above

Subsetting to six planning areas keeps the colour count manageable. The chosen planning areas span the development timeline of Singapore public housing, from mature estates (Bedok, Tampines) to recent towns (Sengkang, Punggol). `alpha = 0.7` mitigates overplotting where points cluster densely.
:::

::: callout-tip
## Reading the plot

Punggol and Sengkang sit visibly closer to the YOUNG vertex than Bedok. This matches the policy intuition that newer Built-To-Order towns attract young families, while mature estates ageing in place skew towards the OLD vertex.
:::

## [13.5.2 Alternative 2: Density Contours]{style="color:  #366092; font-size: 24px;"}

`stat_density_tern()` adds 2D density contours over the triangle, similar to a `geom_density_2d()` overlay on a Cartesian scatter.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  stat_density_tern(geom = "polygon",
                    aes(fill = after_stat(level)),
                    alpha = 0.6) +
  geom_point(size = 0.8, alpha = 0.5) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Density of subzone population structures, 2018",
       subtitle = "Contours reveal the modal age structure across all subzones",
       fill = "Density") +
  theme_rgbw()
```

## Code

```{r}
#| eval: false
ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  stat_density_tern(geom = "polygon",
                    aes(fill = after_stat(level)),
                    alpha = 0.6) +
  geom_point(size = 0.8, alpha = 0.5) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Density of subzone population structures, 2018",
       subtitle = "Contours reveal the modal age structure across all subzones",
       fill = "Density") +
  theme_rgbw()
```
:::

::: callout-note
## Things to learn from the code chunk above

`stat_density_tern()` is the ternary analogue of `stat_density_2d()`. It estimates a 2D density on the triangular surface. Layering `geom_point()` underneath the contours preserves the raw data. `scale_fill_viridis_c(option = "C")` applies the plasma palette.
:::

::: callout-tip
## Reading the plot

The contours highlight a single dense mode somewhere in the ACTIVE-dominant zone, which is the typical Singapore subzone in 2018. Outliers, visible as points outside the densest contours, are the cases worth following up on.
:::

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 13 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: Compositional Cartographer** 🔺

*Press any key to continue...*
:::

`````{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
`````
