---
title: "Hands-on Exercise 9b"
subtitle: "Visualising Geospatial Point Data"
author: "Yee Weicong Mark"
date: "June 15, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    css: styles.css
---

## [22 Visualising Geospatial Point Data]{style="color:  #4682B4; font-size: 38px;"}

## [22.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

Proportional symbol maps, also known as graduated symbol maps, use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon such as counts of people. Like choropleth maps, they come in classed and unclassed versions. The classed ones are known as range-graded or graduated symbols. The unclassed ones are called proportional symbols, where the area of each symbol is proportional to the value of the mapped attribute. This chapter creates a proportional symbol map showing the number of wins by Singapore Pools outlets with the tmap package.

## [22.1.1 Learning Outcome]{style="color:  #366092; font-size: 24px;"}

By the end of this hands-on exercise you will be able to:

1.  import an aspatial data file into R,
2.  convert it into a simple point feature data frame and assign an appropriate projection reference,
3.  plot interactive proportional symbol maps.

## [22.2 Getting Started]{style="color:  #2C5282; font-size: 28px;"}

The tmap package and its related packages are installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## [22.3 Geospatial Data Wrangling]{style="color:  #2C5282; font-size: 28px;"}

## [22.3.1 The Data]{style="color:  #366092; font-size: 24px;"}

The dataset is called `SGPools_svy21`, in csv file format. It has seven columns. The XCOORD and YCOORD columns are the x and y coordinates of Singapore Pools outlets and branches, given in the Singapore SVY21 Projected Coordinate System.

## [22.3.2 Data Import and Preparation]{style="color:  #366092; font-size: 24px;"}

`read_csv()` of readr imports `SGPools_svy21.csv` into R as a tibble data frame called `sgpools`.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

The imported data is examined with `list()`.

```{r}
list(sgpools)
```

::: callout-note
## Things to Learn from the Code Chunk Above

The `sgpools` data is a tibble data frame rather than a common R data frame.
:::

## [22.3.3 Creating an sf Data Frame from an Aspatial Data Frame]{style="color:  #366092; font-size: 24px;"}

`st_as_sf()` of sf converts the `sgpools` data frame into a simple feature data frame.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

::: callout-note
## Things to Learn from the Code Chunk Above

The `coords` argument takes the x-coordinate column name first, then the y-coordinate column name. The `crs` argument takes the coordinate system in EPSG format. EPSG 3414 is the Singapore SVY21 Projected Coordinate System.
:::

The basic information of `sgpools_sf` is displayed with the code chunk below.

```{r}
list(sgpools_sf)
```

::: callout-note
## Things to Learn from the Code Chunk Above

`sgpools_sf` is a point feature class. Its epsg ID is 3414. The bounding box gives the spatial extent of the geospatial data, and a new geometry column has been added.
:::

## [22.4 Drawing Proportional Symbol Map]{style="color:  #2C5282; font-size: 28px;"}

The view mode of tmap creates an interactive proportional symbol map. The code chunk below turns on the interactive mode.

```{r}
tmap_mode("view")
```

## [22.4.1 An Interactive Point Symbol Map]{style="color:  #366092; font-size: 24px;"}

The code chunk below creates an interactive point symbol map.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

## Code

```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```
:::

## [22.4.2 Making It Proportional]{style="color:  #366092; font-size: 24px;"}

A numerical variable is assigned to the size visual attribute to draw a proportional symbol map. The code chunk below assigns the variable Gp1Gp2 Winnings to size.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Code

```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
:::

## [22.4.3 Giving It a Different Colour]{style="color:  #366092; font-size: 24px;"}

The colour visual attribute improves the proportional symbol map further. The code chunk below uses the OUTLET TYPE variable as the colour attribute.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Code

```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
:::

## [22.4.4 Faceted Maps with Synchronised Views]{style="color:  #366092; font-size: 24px;"}

The view mode of tmap also works with faceted plots. The `sync` argument in `tm_facets()` produces multiple maps with synchronised zoom and pan settings.

::: panel-tabset
## Plot

```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

## Code

```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```
:::

::: callout-tip
## Reading the Plot

The two synchronised panels separate the branch outlets from the main outlets while keeping the same zoom and pan. Panning one panel moves the other, so the spatial spread of each outlet type can be compared directly.
:::

The tmap viewer is switched back to plot mode before the session ends.

```{r}
tmap_mode("plot")
```

## [22.5 Reference]{style="color:  #2C5282; font-size: 28px;"}

Kam, T. S. (2025). *R for Visual Analytics*, Chapter 22, Visualising Geospatial Point Data. Singapore Management University.

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 22 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: Point Plotter** 📍

*Press any key to continue...*
:::

```{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
```
