---
title: "Hands-on Exercise 10"
subtitle: "Information Dashboard Design: R Methods"
author: "Yee Weicong Mark"
date: "June 23, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: false
format:
  html:
    css: styles.css
---

## [31 Information Dashboard Design: R Methods]{style="color:  #4682B4; font-size: 38px;"}

## [31.1 Overview]{style="color:  #2C5282; font-size: 28px;"}

An information dashboard packs many small, comparable views into one screen so a reader can scan the state of a business at a glance. This chapter builds the building blocks of such a dashboard in R.

By the end of this hands-on exercise you will be able to:

1.  create a bullet chart using ggplot2,
2.  create sparklines using ggplot2,
3.  build a static dashboard table using gt and gtExtras,
4.  build an interactive dashboard table using reactable and reactablefmtr.

## [31.2 Getting Started]{style="color:  #2C5282; font-size: 28px;"}

The following R packages are used in this exercise.

1.  **tidyverse** for importing, tidying, wrangling and visualising data,
2.  **lubridate** for working with dates and times,
3.  **ggthemes** for additional ggplot2 themes,
4.  **gt** and **gtExtras** for building and styling tables with inline charts,
5.  **reactable** and **reactablefmtr** for building interactive data tables,
6.  **dataui** for the interactive sparklines that reactablefmtr draws,
7.  **svglite** and **ragg** as the graphics backend that gtExtras uses to render its inline sparklines.

The guard below prepares the environment before the packages are loaded. Two of the packages can no longer be installed from CRAN, so they come from their original sources. dataui has always been a GitHub-only package, and reactablefmtr was archived from CRAN in March 2026. reactablefmtr is pure R and its dependencies are still live on CRAN, so no build tools are required. The guard also installs ggplot2 3.5.x into a library folder used only by this page, because gtExtras renders its inline sparklines as black blocks under ggplot2 4.0. The main library keeps ggplot2 4.0 for the other exercises. These steps run only on the first render.

```{r}
options(repos = c(CRAN = "https://cran.r-project.org"))

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
# Render this document against ggplot2 3.5.x from an isolated library folder, so
# the gtExtras sparklines do not render as black blocks. The main library keeps
# ggplot2 4.0 for the other exercises (TH_EX01 needs it for ggstatsplot). This
# affects this page only, because each Quarto file renders in its own session.
gg_lib <- file.path(getwd(), "_ggplot2_35")
dir.create(gg_lib, showWarnings = FALSE)
.libPaths(c(gg_lib, .libPaths()))
if (utils::packageVersion("ggplot2") >= "4.0.0") {
  remotes::install_version("ggplot2", version = "3.5.1",
                           lib = gg_lib, upgrade = "never")
}
if (!requireNamespace("dataui", quietly = TRUE)) {
  remotes::install_github("timelyportfolio/dataui",
                          upgrade = "never")
}
if (!requireNamespace("reactablefmtr", quietly = TRUE)) {
  remotes::install_version("reactablefmtr",
                           version = "2.0.0",
                           upgrade = "never")
}
if (!requireNamespace("svglite", quietly = TRUE)) install.packages("svglite")
if (!requireNamespace("ragg", quietly = TRUE)) install.packages("ragg")
```

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               gt, gtExtras, svglite, ragg, tidyverse)
library(reactablefmtr)
library(dataui)
```

## [31.3 Importing Microsoft Access Database]{style="color:  #2C5282; font-size: 28px;"}

## [31.3.1 The Data Set]{style="color:  #366092; font-size: 24px;"}

A personal database in Microsoft Access mdb format called `Coffee Chain` is used. It records sales and budgeted sales for a chain of coffee products.

## [31.3.2 Importing Database into R]{style="color:  #366092; font-size: 24px;"}

`odbcConnectAccess2007()` of the RODBC package imports a database query table into R. The code chunk below is shown for reference. It is not executed here because the Access driver runs on a 32-bit R session rather than the 64-bit session used to render this page.

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/rds/CoffeeChain.rds")
odbcClose(con)
```

::: callout-note
## Things to Learn from the Code Chunk Above

`odbcConnectAccess2007()` opens a connection to the Access file, `sqlFetch()` pulls a named query into an R data frame, and `odbcClose()` closes the connection. The result is cached as an rds file so the rest of the exercise can read it directly without the Access driver.
:::

## [31.3.3 Data Preparation]{style="color:  #366092; font-size: 24px;"}

The prepared `CoffeeChain.rds` is read into R. This step is the practical starting point once the rds file is available.

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")
```

Sales and budgeted sales are aggregated at the product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## [31.3.4 Bullet Chart in ggplot2]{style="color:  #366092; font-size: 24px;"}

A bullet chart shows a single measure against qualitative ranges and a target marker. The code chunk below builds one from layered `geom_col()` bars and a `geom_errorbar()` target line.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current),
           width=0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax= target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## Code

```{r}
#| eval: false
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current),
           width=0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax= target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

The three wide grey bars run from light to dark to mark qualitative bands of performance. The thin black bar is the actual value for each product, and the red `geom_errorbar()` marks the target. `coord_flip()` turns the chart on its side so the product labels are readable.
:::

::: callout-tip
## Reading the Plot

A black bar that reaches past the red marker means the product met its target. The grey bands give quick context for whether a value is poor, fair or good without reading the axis precisely.
:::

## [31.4 Plotting Sparklines Using ggplot2]{style="color:  #2C5282; font-size: 28px;"}

A sparkline is a small line chart drawn without axes that shows the shape of a series in the space of a word. This section builds one sparkline per product from the monthly sales.

## [31.4.1 Preparing the Data]{style="color:  #366092; font-size: 24px;"}

Monthly sales per product are derived from the daily records.

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

The minimum, maximum and end-of-series sales per product are computed for the labelled points.

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))
```

The 25th and 75th percentiles per product are computed for the shaded band.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales,
                              0.25),
            quart2 = quantile(Sales,
                              0.75)) %>%
  right_join(sales_report)
```

## [31.4.2 Sparklines in ggplot2]{style="color:  #366092; font-size: 24px;"}

The sparklines are faceted one row per product, with a trend line, an interquartile band and labelled extreme points.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 8
ggplot(sales_report, aes(x=Month, y=Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, ymax = quart2),
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

## Code

```{r}
#| eval: false
ggplot(sales_report, aes(x=Month, y=Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, ymax = quart2),
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

`facet_grid(Product ~ ., scales = "free_y")` stacks one small panel per product and lets each panel scale its own y-axis. `geom_ribbon()` shades the interquartile band, the red and blue points mark the lowest and highest months, and `theme_tufte()` strips the axes so only the shape of the series remains.
:::

::: callout-tip
## Reading the Plot

Each row reads as a single glance at one product's year. The blue and red points anchor the peak and trough, and the grey band shows the middle half of the values, so an unusual month stands out against it.
:::

## [31.5 Static Information Dashboard Design: gt and gtExtras Methods]{style="color:  #2C5282; font-size: 28px;"}

The gt package builds publication-quality tables, and gtExtras adds helper functions that draw small charts inside table cells. This section places a bullet chart inside a table.

## [31.5.1 Plotting a Simple Bullet Chart]{style="color:  #366092; font-size: 24px;"}

`gt_plt_bullet()` draws an inline bullet chart for each row, comparing the current value against its target.

::: panel-tabset
## Plot

```{r}
#| echo: false
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
              target = target,
              width = 60,
              palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## Code

```{r}
#| eval: false
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
              target = target,
              width = 60,
              palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

`gt::gt()` turns the data frame into a gt table. `gt_plt_bullet()` replaces the numeric column with an inline bullet chart, where `column` is the actual value and `target` is the reference. `gt_theme_538()` applies a clean newsroom-style theme.
:::

## [31.6 Sparklines: gtExtras Method]{style="color:  #2C5282; font-size: 28px;"}

gtExtras can also draw sparklines inside a table. The data is first reshaped into the form gtExtras expects.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date,
                        label = TRUE,
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

gtExtras functions almost always require a data frame with list columns. The code chunk below collapses each product's monthly sales into a single list column.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

::: callout-note
## Things to Learn from the Code Chunk Above

`list(Sales)` packs the twelve monthly values of each product into one cell. A list column holds a vector in every row, which is the shape `gt_plt_sparkline()` reads to draw one sparkline per product.
:::

## [31.6.1 Plotting Coffee Chain Sales Report]{style="color:  #366092; font-size: 24px;"}

`gt_plt_sparkline()` draws an inline sparkline from the list column.

::: panel-tabset
## Plot

```{r}
#| echo: false
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

## Code

```{r}
#| eval: false
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

`same_limit = FALSE` lets each sparkline use its own y-scale, so the shape of every product is visible even when the products differ greatly in sales volume.
:::

## [31.6.2 Adding Statistics]{style="color:  #366092; font-size: 24px;"}

Summary statistics per product are computed and shown in a gt table.

::: panel-tabset
## Plot

```{r}
#| echo: false
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

## Code

```{r}
#| eval: false
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```
:::

::: callout-note
## Things to Learn from the Code Chunk Above

`fmt_number(columns = 4, decimals = 2)` formats the Average column to two decimal places. The Min and Max columns are left as whole numbers.
:::

## [31.6.3 Combining the data.frame]{style="color:  #366092; font-size: 24px;"}

The sparkline list column and the statistics are built separately, then joined into one table.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

```{r}
sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

## [31.6.4 Plotting the Updated data.table]{style="color:  #366092; font-size: 24px;"}

The combined table shows the statistics alongside the sparkline.

::: panel-tabset
## Plot

```{r}
#| echo: false
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Code

```{r}
#| eval: false
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```
:::

## [31.6.5 Combining Bullet Chart and Sparklines]{style="color:  #366092; font-size: 24px;"}

A bullet chart and a sparkline can sit in the same table. The actual and target values are prepared, joined onto the table, then both inline charts are drawn.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

::: panel-tabset
## Plot

```{r}
#| echo: false
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## Code

```{r}
#| eval: false
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```
:::

::: callout-tip
## Reading the Plot

Each row now carries the year's shape as a sparkline and the target comparison as a bullet chart, alongside the summary numbers. The table reads as a compact dashboard where every product can be judged on trend and on target in one line.
:::

## [31.7 Interactive Information Dashboard Design: reactable and reactablefmtr Methods]{style="color:  #2C5282; font-size: 28px;"}

reactable builds interactive tables, and reactablefmtr adds interactive sparklines. The interactive sparklines are drawn by the dataui package, which was installed and loaded in section 31.2 so that reactablefmtr could register its sparkline functions.

## [31.7.1 Plotting Interactive Sparklines]{style="color:  #366092; font-size: 24px;"}

As with gtExtras, the monthly sales are packed into a list column first.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

`react_sparkline()` then draws the interactive sparkline inside the reactable.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```
:::

::: callout-tip
## Reading the Plot

Hovering over a sparkline reveals the value at each month, which the static versions cannot show. The interactivity turns the small chart into a quick lookup tool.
:::

## [31.7.2 Changing the Pagesize]{style="color:  #366092; font-size: 24px;"}

The default page size is ten rows. `defaultPageSize` changes it so all thirteen products show on one page.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```
:::

## [31.7.3 Adding Points and Labels]{style="color:  #366092; font-size: 24px;"}

`highlight_points` marks the minimum and maximum values, and `labels` prints the first and last values.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```
:::

## [31.7.4 Adding Reference Line]{style="color:  #366092; font-size: 24px;"}

`statline` adds a reference line at a summary statistic such as the mean.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```
:::

## [31.7.5 Adding Bandline]{style="color:  #366092; font-size: 24px;"}

A bandline shades a range behind the sparkline. The code chunk below adds the interquartile band.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```
:::

## [31.7.6 Changing from Sparkline to Sparkbar]{style="color:  #366092; font-size: 24px;"}

`react_sparkbar()` shows the values as bars instead of a line.

::: panel-tabset
## Plot

```{r}
#| echo: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

## Code

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
:::

::: callout-tip
## Reading the Plot

The sparkbar suits readers who compare individual months more than the overall trend. The bandline and mean line give the same context as the sparkline version while each month reads as a discrete bar.
:::

## [31.8 Reference]{style="color:  #2C5282; font-size: 28px;"}

Kam, T. S. (2026). *R for Visual Analytics*, Chapter 31, Information Dashboard Design: R Methods. Singapore Management University.

::: {style="text-align: center; padding: 10px 30px; background: #0a0e27; color: #00d4ff; border: 4px solid #00d4ff; border-radius: 10px; margin-top: 60px; font-family: 'Courier New', monospace; text-shadow: 0 0 10px #00d4ff; animation: pulse 2s infinite;"}
# 🕹️ LEVEL COMPLETE 🕹️

## ★ ★ ★ ★ ★

### CHAPTER 31 CLEARED!

**+1000 XP** · **ACHIEVEMENT UNLOCKED: Dashboard Architect** 📊

*Press any key to continue...*
:::

```{=html}
<style>
@keyframes pulse {
  0%, 100% { box-shadow: 0 0 30px rgba(0, 212, 255, 0.5); }
  50% { box-shadow: 0 0 50px rgba(0, 212, 255, 0.9); }
}
</style>
```
